---
title: "Assignment 2: Coding Basics"
author: Molly Bruce
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.

## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 
2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. The sequence command generates a list of numbers beginning at the first number in the parenthesis, ending at the second number in the parenthesis, and by increments of the third number in the parenthesis. 

S = seq(1, 100, 4)

#2. These two functions are fairly self-explanatory--include in the parenthesis the variable you've assigned to the group of numbers you'd like to analyze. Because I know I'll be running a comparative analysis of the mean and median of our sequence, I've also assigned the mean and median to two new variables.

M = median(S)
A = mean(S)

#3. I chose to use an if-else statement to determine how the mean and median compare. It's important to place the tildes, print statements, and else statement in particular locations (as shown below), otherwise R will throw back an error message. The outcome of this print statement is populated in the Console tab.

if (A > M) {
  print("mean is greater than median")
  } else {
  print("mean is not greater than median")
  }

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}

#5. I have assumed that, by telling us to create these vectors, you'd also like us to populate them with information. 

Name <- c("Aaron", "Beth", "Charlie", "Debbie")
Grade <- c(84, 79, 48, 91)
PassFail <- c(TRUE, TRUE, FALSE, TRUE)

#6. Though the important function is the class() function, I've also included descriptive print statements above each class() function.

print("The vector titled Name is vector type: ") 
  class(Name)
print("The vector titled Grade is vector type: ") 
  class(Grade)
print("The vector titled PassFail is vector type: ") 
  class(PassFail)

#7. I used the cbind function to assign the 3 individual vectors to a signle dataframe titled StudentPerformanceDF (DF being short for dataframe). I printed that dataframe for my own benefit.

StudentPerformanceDF <- cbind(Name,Grade,PassFail)
print(StudentPerformanceDF)

#8.Though I already gave my vectors fairly logical and informative titles at #5 and these titles became column headings when the vectors were combined into a single dataframe, I've shown the process for assigning new (and hopefully more descriptive) column titles within a dataframe using the data.frame() function and then printed the updated dataframe for my own benefit.

colnames(StudentPerformanceDF)
StudentPerformanceDF_titled <- data.frame("StudentName"=Name,"TestScore"=Grade,"StudentPassed"=PassFail)
print(StudentPerformanceDF_titled)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A dataframe can have data of different types, while a matrix must have uniform datatypes. So, for instance, we have a character field, a numeric field, and a logical field in our dataframe. However, a matrix could only have one of these--perhaps only numeric values. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

#10. Below, I am simply creating an in/else function. As written, this function would only apply to the first Grade in our vector; it would not iterate through all students' grades. We will address this issue next. 

BadPassFailFunction = if (Grade >= 50) {
  print("Passed")
  } else {
  print("Failed")
  }

#11. Next, I use the ifelse approach which allows me to iterate through all values in a vector and output an answer for each value. 

GoodPassFailFunction = ifelse(Grade >=50, "Passed", "Failed")

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
