---
title: "Assignment 4: Data Wrangling"
author: Molly Bruce
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, Feb 16 @ 11:59pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1
getwd() 
# Commented out for purposes of knitting the file
#setwd("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021")

# Commented out for purposes of knitting the file
#install.packages(tidyverse)
#install.packages(lubridate)
library(tidyverse)
library(lubridate)

#The ideal format for calling these datafiles is a relative path using the format
#EPAair_O3_NC_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
#however, this method has thrown errors for me on Assignment 3 and also on this assignment.
#Therefore, I coded the enture path even though this is less resilient and less prefered. 
EPAair_O3_NC_2019 <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Raw/EPAair_O3_NC2019_raw.csv")
EPAair_O3_NC_2018 <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Raw/EPAair_O3_NC2018_raw.csv")
EPAair_PM25_NC_2019 <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Raw/EPAair_PM25_NC2019_raw.csv")
EPAair_PM25_NC_2018 <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Raw/EPAair_PM25_NC2018_raw.csv")

#2
#Exploring the Ozone 2019 dataset 
colnames(EPAair_O3_NC_2019)
head(EPAair_O3_NC_2019)
summary(EPAair_O3_NC_2019)
str(EPAair_O3_NC_2019)
dim(EPAair_O3_NC_2019)

#Exploring the Ozone 2018 dataset 
colnames(EPAair_O3_NC_2018)
head(EPAair_O3_NC_2018)
summary(EPAair_O3_NC_2018)
str(EPAair_O3_NC_2018)
dim(EPAair_O3_NC_2018)

#Exploring the PM25 2019 dataset 
colnames(EPAair_PM25_NC_2019)
head(EPAair_PM25_NC_2019)
summary(EPAair_PM25_NC_2019)
str(EPAair_PM25_NC_2019)
dim(EPAair_PM25_NC_2019)

#Exploring the PM25 2018 dataset 
colnames(EPAair_PM25_NC_2018)
head(EPAair_PM25_NC_2018)
summary(EPAair_PM25_NC_2018)
str(EPAair_PM25_NC_2018)
dim(EPAair_PM25_NC_2018)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 I am assuming that by saying "Change date to date," 
#you want us to parse the string turned factor into a traditional date object 
EPAair_O3_NC_2019$Date <- as.Date(EPAair_O3_NC_2019$Date, format = "%m/%d/%Y")
EPAair_O3_NC_2018$Date <- as.Date(EPAair_O3_NC_2018$Date, format = "%m/%d/%Y")
EPAair_PM25_NC_2019$Date <- as.Date(EPAair_PM25_NC_2019$Date, format = "%m/%d/%Y")
EPAair_PM25_NC_2018$Date <- as.Date(EPAair_PM25_NC_2018$Date, format = "%m/%d/%Y")

#4 
EPAair_O3_NC_2019_selected <- 
  select(EPAair_O3_NC_2019, Date, DAILY_AQI_VALUE, Site.Name, 
         AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPAair_O3_NC_2018_selected <- 
  select(EPAair_O3_NC_2018, Date, DAILY_AQI_VALUE, Site.Name, 
         AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPAair_PM25_NC_2019_selected <- 
  select(EPAair_PM25_NC_2019, Date, DAILY_AQI_VALUE, Site.Name, 
         AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPAair_PM25_NC_2018_selected <- 
  select(EPAair_PM25_NC_2018, Date, DAILY_AQI_VALUE, Site.Name, 
         AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5
EPAair_PM25_NC_2019_selected$AQS_PARAMETER_DESC <- "PM2.5"
EPAair_PM25_NC_2018_selected$AQS_PARAMETER_DESC <- "PM2.5"

#6
write.csv(EPAair_O3_NC_2019_selected,"C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_O3_NC2019_processed.csv", row.names = FALSE)
write.csv(EPAair_O3_NC_2018_selected,"C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_O3_NC2018_processed.csv", row.names = FALSE)
write.csv(EPAair_PM25_NC_2019_selected,"C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_PM25_NC2019_processed.csv", row.names = FALSE)
write.csv(EPAair_PM25_NC_2018_selected,"C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_PM25_NC2018_processed.csv", row.names = FALSE)
```

## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7
# I am combining the four datasets with identical column names
EPAair_NC <- rbind(EPAair_O3_NC_2019_selected, 
                   EPAair_O3_NC_2018_selected, 
                   EPAair_PM25_NC_2019_selected, 
                   EPAair_PM25_NC_2018_selected)

#8
# I am assuming that you want us to use the intersect function to find the sites
# the four data frames have in common, as opposed to filtering based on the list 
# you provided, though either method would work.
EPAair_NC_piped <-
  EPAair_NC %>%
  # Finding the sites the four dataframes have in common using intersect
  filter(Site.Name %in% c(intersect(
    intersect(EPAair_O3_NC_2019_selected$Site.Name,
              EPAair_O3_NC_2018_selected$Site.Name),
    intersect(EPAair_PM25_NC_2019_selected$Site.Name,
              EPAair_PM25_NC_2018_selected$Site.Name))) & Site.Name != "") %>%
  # Split-Apply-Combine 
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>% 
  summarise(mean.AQI = mean(DAILY_AQI_VALUE), 
            mean.lat = mean(SITE_LATITUDE), 
            mean.long = mean(SITE_LONGITUDE), .groups = "keep") %>%
  # Adding the Month and Year columns 
  mutate(Month = month(Date), Year = year(Date)) 

# Throwing this into a csv because I worked really hard and was worried I'd mess it up
write.csv(EPAair_NC_piped, row.names = FALSE, 
          file ="C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_NC_piped.csv")

#9 
# Pivoting the dataframe 
EPAair_NC_piped_spread <- 
  EPAair_NC_piped %>% 
  pivot_wider(id_cols = c(Date, Month, Year, Site.Name, COUNTY, mean.lat, mean.long),
                names_from = AQS_PARAMETER_DESC, values_from = mean.AQI)

#10 
dim(EPAair_NC_piped_spread)

#11 
write.csv(EPAair_NC_piped_spread, row.names = FALSE, 
          file ="C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

```

## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a
# Split-Apply-Combine in a pipe again 
EPAair_NC_piped_processed <- 
  EPAair_NC_piped_spread %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(mean.AQI.PM2.5 = mean(PM2.5), 
            mean.AQI.Ozone = mean(Ozone))

#12b
# Using a separate pipe for this drop_na approach, though I could have done it in the 12a pipe
EPAair_NC_piped_processed_clean <-
  EPAair_NC_piped_processed %>%
  drop_na(c(Month, Year))

#13
dim(EPAair_NC_piped_processed_clean)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: na.omit is general; it removes rows with NA values in ANY column, even if you try to specify a particular column. Meanwhile, drop_na is targeted; it allows users to specify the column where it should search and then drops rows where the NAs are present in that specified column. In our case, running the na.omit function returns a dataframe with dimensions 101 by 5, meaning that, as predicted, it removed a lot more rows because it also tosses rows with NAs in the Mean.AQI.PM2.5 and Mean.AQI.Ozone columns. 
