---
title: "Assignment 5: Data Visualization"
author: Molly Bruce 
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 23 at 11:59 pm.

## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 

getwd()
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("cowplot")
library(tidyverse)
library(ggplot2)
library(cowplot)

NTL_LTER_nutrients_processed <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NTL_LTER_gathered_processed <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)

#2

# Verified in Environment tab that the sampledate column is read in as factor
NTL_LTER_nutrients_processed$sampledate <- as.Date(NTL_LTER_nutrients_processed$sampledate, format = "%Y-%m-%d")
NTL_LTER_gathered_processed$sampledate <- as.Date(NTL_LTER_gathered_processed$sampledate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") #alternative: legend.position + legend.justification

theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
NTL_LTER_Scatter <- ggplot(NTL_LTER_nutrients_processed[NTL_LTER_nutrients_processed$po4 <=50,], 
                         aes(x = tp_ug, 
                             y = po4, 
                             color = lakename)) + 
  geom_point(size = .9, alpha = .5) + 
  geom_smooth(method = lm, se = FALSE, color = "black") +
  ylab("Phosphate") + 
  xlab("Phosphorus") +
  mytheme
print(NTL_LTER_Scatter)
```

5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
NTL_LTER_Box_Temp <- ggplot(NTL_LTER_nutrients_processed, 
                            aes(x = as.factor(month), 
                                y = temperature_C)) + 
  geom_boxplot(aes(color = lakename)) + 
  ylab("Temperature (C)") + 
  xlab("Month") + 
  scale_x_discrete(month)
print(NTL_LTER_Box_Temp)
  
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

```{r}

```

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
