---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: Molly Bruce
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1

getwd()

library(agricolae)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scales)
library(lubridate)

# Again, I'm providing full paths because 
# RStudio on my virtual machine won't process the relative paths
# I recognize that relative paths are preferable

LakeChem_raw <- 
  read.csv("C:/Users/mmb88/Desktop/Environmental_Data_Analytics_2021/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
           stringsAsFactors = TRUE) 

# Verified before this in Environment tab that sampledate column is read as factor
LakeChem_raw$sampledate <- 
  as.Date(LakeChem_raw$sampledate, format = "%m/%d/%y")

#2

# I could add more to this theme but will leave it more simple for now 
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right", 
        panel.grid.major = element_line(colour = "gray")) 
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: July mean lake temperature in any lake does not change with depth. 
Ha: July mean lake temperature in all lakes changes with depth. 

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

# I am making a month column so that I can filter by this in my pipe
LakeChem_raw <- mutate(LakeChem_raw, month = month(sampledate)) 

LakeChem_piped <- 
  LakeChem_raw %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit

#5

LakeChem_scatterTD <- ggplot(LakeChem_piped,
                aes(x = depth, 
                    y = temperature_C)) +
  geom_point(size = .8, alpha = .5, color = "dodgerblue") + 
  geom_smooth(method = lm, se = FALSE, color = "black") +
  labs(title = "July Lake Temperature By Depth",
       y = "Temperature (C)",
       x = "Depth Below Surface (m)") + 
  theme(plot.title = element_text(hjust=0.5)) +
  scale_y_continuous(limits = c(0, 35), expand = c(0, 0)) 
print(LakeChem_scatterTD)

```

6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This scatter plot and corresponding line of best fit suggest that temperature tends to decrease with depth (an inverse relationship). However, given the long tail on the righthand side of the plot, it appears as though there's a floor for temperature, even while depth continues to increase. Conceptually, this makes sense; deep portions of lakes do not freeze first so we wouldn't expect temperatures to continue to drop below 0 as depth increases. 

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

LakeChem_lr <- lm(data=LakeChem_raw, temperature_C ~ depth)
summary(LakeChem_lr)

```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to this linear regression, 67.42% of temperature variability is explained by depth. There are 34,754 degrees of freedom. The results are statistically significant based on the p-value which is nearly 0. The temperature is predicted to decrease roughly 1.7 degrees for each meter increase in depth. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 

9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

AIC <- lm(data = LakeChem_piped, temperature_C ~ year4 + daynum + depth)
step(AIC)

#10

#multi_regression <- lm(data = LakeChem_piped, 
#                       temperature_C ~ year4 + daynum + depth)
# Because we are looking at the same variables, I'm running summary on AIC. 
# However, I could also run summary on the above hashed out multi_regression 
summary(AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC suggests that all three variables--year, day, and depth--are statistically important for predicting temperature. However, it suggests that year is less important than day and depth. These three variables in tandem account for 74.11% of the observed variance in temperature. By comparison, the depth variable on its own only accounts for 67.42% of the observed variance. As such, it is an improvement over the model with only depth.  

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

ANOVA <- aov(data = LakeChem_piped, temperature_C ~ lakename)
summary(ANOVA)

linear_model <- lm(data = LakeChem_piped, temperature_C ~ lakename)
summary(linear_model)

#plot(linear_model)

``` 

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is. We see some fairly large T-values (in absolutele value) indicating that it is statistically unlikely that temperature does NOT vary by lake. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.height=4, fig.width=6}
#14.

LakeChem_scatter_lakes <- ggplot(LakeChem_piped,
                aes(x = depth, 
                    y = temperature_C, 
                    color = lakename)) +
  geom_point(size = .8, alpha = .5) + 
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "July Lake Temperature By Depth",
       y = "Temperature (C)",
       x = "Depth Below Surface (m)", 
       color = "Lake Name") + 
  theme(plot.title = element_text(hjust=0.5)) +
  scale_y_continuous(limits = c(0, 35), expand = c(0, 0)) 
print(LakeChem_scatter_lakes)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}

#15 TukeyHSD() computes Tukey Honest Significant Differences

TukeyHSD(ANOVA)
Tukey <- HSD.test(ANOVA, "lakename", group = TRUE)
Tukey

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically, Paul Lake, Peter Lake, and Ward Lake have the same temperature (all group c). No lake(s) have mean temperatures that are statistically distinct from the other lakes. 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could also run the Two-Sample T-Test. 
